A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-13 jour/jour.tex
more

Given a fixed signature~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
$\thD$-satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between $\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as a tableau-like calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. The presentation is
inspired by Barrett et al.\ \cite{barrett-et-al-2007} but
higher-level, using unoriented equations instead of oriented ones.

@@

Our calculus for $\thD$ consists of derivation rules.
A derivation rule can be applied to $\Ec$ if %all
the specified
preconditions are met.
The conclusion either specifies equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
One rule has multiple conclusions, % separated by $\ror$,
denoting branching.
%
An application of a rule is \emph{redundant} if one of its non-$\bot$
conclusions leaves $\Ec$ unchanged.
A \emph{derivation tree} is a %finite
tree whose nodes are finite sets of
equalities, such that child nodes are obtained by a nonredundant application of a
derivation rule to the parent. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

\paragraph{Phase 2: Applying Acyclicity and Uniqueness \rm(Fig.~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
%The premises of the rules that make up the second phase refer to
We describe the rules in this phase in terms of a mapping $\Val$
that associates with each equivalence class a $\mu$-term as its representative.
%\rem{Not explicitly constructed, add here?}

@@

The mapping $\Val$ is defined as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$. %not occurring in $\tEc$,
For a term $t \in \tEc$, 
we write $\Varec{t}$ to denote the variable associated with
the equivalence class $\ec{t}$.
Initially, we set $\Val\ec{u} := {\Varec{u}}$ for each equivalence class $\ec{u}$. Because $\Varec{u}$ is unspecified,
this indicates that there are initially no
constraints on the values for any equivalence class $\ec{u}$. 
The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

>>>>

Given a fixed signature~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
$\thD$-satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between $\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as a tableau-like calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. The presentation is
inspired by Barrett et al.\ \cite{barrett-et-al-2007} but at
a higher level, using unoriented equations instead of oriented ones.

@@

Our calculus for $\thD$ consists of derivation rules.
A derivation rule can be applied to $\Ec$ if %all
the specified
preconditions are met.
The conclusion either specifies equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
One of the rules has multiple conclusions, % separated by $\ror$,
denoting branching.
%
An application of a rule is \emph{redundant} if one of its non-$\bot$
conclusions leaves $\Ec$ unchanged.
A \emph{derivation tree} is a %finite
tree whose nodes are finite sets of
equalities, such that child nodes are obtained by a nonredundant application of a
derivation rule to the parent. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

\paragraph{Phase 2: Applying Acyclicity and Uniqueness \rm(Fig.~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
%The premises of the rules that make up the second phase refer to
The rules in this phase are described in terms of a mapping $\Val$
that assigns to each equivalence class a $\mu$-term as its representative.
%\rem{Not explicitly constructed, add here?}

@@

The mapping $\Val$ is defined as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$. %not occurring in $\tEc$,
For a term $t \in \tEc$, 
we write $\Varec{t}$ to denote the variable associated with
the equivalence class $\ec{t}$.
Initially, we set $\Val\ec{u} := {\Varec{u}}$ for each equivalence class $\ec{u}$. Because $\Varec{u}$ is unspecified,
this indicates that there are initially no
constraints on the values for any equivalence class $\ec{u}$. 
The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the variable-\emph{capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.
